```{r setup, include=FALSE}

knitr::opts_chunk$set(
  echo = TRUE,
  eval = TRUE,
  message=FALSE, 
  warning=FALSE, 
  fig.width = 10,
  fig.align = "center"
  )

library( DWVpack )
library( tidyverse )
library( dplyr )

```

# Tidy Data {.unnumbered}

Have you ever opened a dataset and immediately thought, “This is a mess”?
Columns don’t make sense.
Dates are inconsistent.
And even if you can read it, it’s hard to actually do anything with it.
If so, then your second thought might have been, "this is going to be a lot of work to fix".

Sound familiar?
That’s not a coincidence.
Rather it is the reality of working with open-source data or data "in the wild".
In data wrangling, we want to corral that wild data and get it into a format with which we can work.

There’s a common misconception that data analysis is mostly about: building models, creating visualizations, creating output, etc.
But in practice, most analysts spend the majority of their time doing something cleaning and organizing data.
Ironically, when we (i.e. scholars) teach methods and statistics for crime analysis, most of the time is spent on analysis and very little time is spent on preparing data.

"Tidying" data is one of the most important skills you can develop.
In this chapter we are going to focus on turning messy data into something you can actually analyze.

## What Is “Tidy Data”?

We’re going to use a concept from the [tidyverse](https://tidyverse.org/)(a collection of R packages for data transformation) called *tidy data*.
This gives us a standard way to structure datasets so they are easy to work with.
Also, organizing data in this way is quickly becoming an industry standard, so knowing what it is and how to do it is valuable.

A dataset is considered tidy when it follows three simple rules:

1.  Each variable has its own column
2.  Each observation has its own row
3.  Each value lives in a single cell

That might sound simple and obvious, but most "wild" datasets violate at least one of these rules and are therefore "untidy".
Let's see some examples.

### Untidy Data: Example 1

```{r, echo=FALSE}

bad_data_1 <- tibble(
  incident_id = 1:4,
  demographics = c("Male_Adult", "Female_Juvenile", "Male_Juvenile", "Female_Adult")
)

bad_data_1

```

See any issues?
You probably noticed that “gender” and “age group” are mashed together in the variable `demographics`.
This is a problem because `demographics` is actually two variables.
How might we solve this problem?

Well, we just need to create two variables:

```{r, echo=FALSE}

good_data_1 <- bad_data_1 |>
  separate(demographics, into = c("gender", "age_group"), sep = "_")

good_data_1

```

### Untidy Data: Example 2

Try this one:

```{r, echo=FALSE}

bad_data_2 <- tibble(
  precinct = c("A", "B"),
  arrests_2022 = c(50, 30),
  arrests_2023 = c(60, 45)
)

bad_data_2

```

If you said “year” is hidden in column names, then you win a prize!
Data organized in this way make it hard to analyze trends by year.
Rather, we would prefer data like this:

```{r, echo=FALSE}

good_data_2 <- bad_data_2 |>
  pivot_longer(
    cols = starts_with("arrests"),
    names_to = "year",
    values_to = "arrests"
  ) |> 
  mutate(year = str_extract(year, "\\d{4}"))

good_data_2

```

Note the difference.
We know have a variable for "year".

### Untidy Data: Example 3

Last one (you got this!):

```{r, echo=FALSE}

bad_data_3 <- tibble(
  incident_id = 1:2,
  subjects = c("John;Mike", "Sara;Anna")
)

bad_data_3

```

These data are untidy because two people are stored in one row.
It would be difficult to count or analyze individuals properly.
We can fix this by making a variable for "subjects" and listing each one:

```{r, echo=FALSE}

good_data_3 <- bad_data_3 |>
  separate_rows(subjects, sep = ";")

good_data_3

```

### Why Tidy?

Why ensure that your data is tidy?
There are two main advantages:

1.  There’s a general advantage to picking one consistent way of storing data. If you have a consistent data structure, it’s easier to learn the tools that work with it because they have an underlying uniformity. 2, There’s a specific advantage to placing variables in columns because it allows R’s vectorized nature to shine. Most built-in R functions work with vectors of values. That makes transforming tidy data feel particularly natural.

`dplyr` (remember `dplyr`!?!), `ggplot2`, and all the other packages in the `tidyverse` are designed to work with tidy data.

Since we will be working with functions that tidy data, we need to install `tidyverse`:

```{r, eval=FALSE}
install.packages( "tidyverse" )
```

After installing (or if you already have the package installed), we need to use the `library()` function to load it:

```{r, eval=FALSE}
library( tidyverse )
```

<br>

## Reshaping Data

Now that we know what tidy data should look like, the next step is learning how to transform messy data into that structure.
One of the most common problems you’ll run into is that your data is in the wrong *shape*.

Take a look at this example:

```{r, echo=FALSE}

data_wide <- tibble(
  precinct = c("A", "B"),
  arrests_2022 = c(50, 30),
  arrests_2023 = c(60, 45)
)

data_wide

```

If I asked you to give me a table of "arrests" over time, what problems do you see?

You probably noticed that the year is embedded in column names.
This makes it had to extract this information for year.

### Wide vs Long Data

This issue we saw in the example above has a specific name.
These data are **wide**, meaning that there are many columns and the variable (in this case "year") is spread over multiple columns.
This is in contrast to data that is **long**, where there are fewer columns and values stacked into rows

Think of it like this:\
- “Wide data spreads information across columns.” - “Long data stacks information down rows.”

#### `pivot_longer`

There is where the `pivot_longer` function comes in.

First, let's create our simple **wide** data object:

```{r}

data_wide <- data.frame(
  precinct = c( "A", "B" ),
  arrests_2022 = c( 50, 30 ),
  arrests_2023 = c( 60, 45 )
)

data_wide

```

Now, let's create the **long** data object.
Let me show you what it does first and we will go through each line:

```{r}

data_long <- data_wide |>
  pivot_longer(
    cols = c( "arrests_2022", "arrests_2023"),
    names_to = "year",
    values_to = "arrests"
  )

data_long

```

Ok, let's see what we did:

-   `cols = c( "arrests_2022", "arrests_2023")` asks "which column are we resphaping?" and we use two column names (note that we have to use `c()` to combine them into a vector.

-   `names_to = "year"` tells `pivot_longer()` where the old column names are supposed to go.

-   `values_to = "arrests"` tells the function what to do with the values.

In this case, they are going to go to a variable called `num_arrests`.

If you are a bit lost, think about it like this: we are taking column names and turning them into data.

If you are a visual learner, think of it this way:

```{mermaid}
flowchart LR

A["WIDE DATA<br/>(columns contain values)<br/><pre>
┌───────────┬───────────────┬───────────────┐
│ precinct  │ arrests_2022  │ arrests_2023  │
├───────────┼───────────────┼───────────────┤
│ A         │ 50            │ 30            │
│ B         │ 60            │ 45            │
└───────────┴───────────────┴───────────────┘
</pre>"]

B["pivot_longer()<br/>
cols = starts_with('arrests')<br/>
names_to = 'year'<br/>
values_to = 'arrests'"]

C["LONG DATA<br/>(values stacked into rows)<br/><pre>
┌───────────┬────────────────┬─────────┐
│ precinct  │ year           │ arrests │
├───────────┼────────────────┼─────────┤
│ A         │ arrests_2022   │ 50      │
│ A         │ arrests_2023   │ 60      │
│ B         │ arrests_2022   │ 30      │
│ B         │ arrests_2023   │ 45      │
└───────────┴────────────────┴─────────┘
</pre>"]

A --> B --> C
```

We are almost to a tidy dataset.
What is missing?
Take a look at the values for the `year` variable.
It contains the string "arrests\_" which was carried over from our transformation.
What we need to is remove this information.
There is a nice function called `str_remove()` which takes an object or column and removes any characters.
The `str_remove()` function is in a package called `stringr` which is part of the `tidyverse`.
We can put that into the `mutate()` function we we already saw in `dpylr`.

```{r}

# make sure we load the functions
library( dplyr )

# now remove the unncessary text
data_long <- data_long |>
  mutate(
    year = str_remove(year, "arrests_")
  )

```

Now, we can create a table using the tools from `dplyr`!

Perhaps average arrests by year:

```{r}

data_long |> 
  group_by( year ) |> 
  summarize( avg_arrests = mean( arrests ) )

```

Or average arrests by precinct:

```{r}

data_long |> 
  group_by( precinct ) |> 
  summarize( avg_arrests = mean( arrests ) )
  
```

### `pivot_wide`

So far, we’ve taken messy wide data and made it tidy.
But sometimes we actually want to go the other direction.
This usually happens when we’re preparing results for reports or summaries.
There is where the `pivot_wide` function comes in.
First, let's create a simple **long** data object.
Suppose we had some data like this for crimes in two neighborhoods:

```{r}

data_long <- tibble(
  neighborhood = c("A", "A", "A", "B", "B", "B" ),
  year = c( "2020", "2021", "2022","2020", "2021", "2022" ),
  crimes = c(50, 60, 30, 45, 90, 40 )
)

data_long

```

If we want this in a format that’s easier to read in a report we could have one row per precinct and separate columns for each year.
We can do that by creating a data object that is in **wide** format using the `pivot_wide()` function:

```{r}

data_wide <- data_long |>
  pivot_wider(
    names_from = year,
    values_from = crimes
  )

data_wide

```

Ok, let's see what we did:

-   `names_from = "year"` tells `pivot_wider()` to use values in the year column as new column names.

-   `values_from = "crimes"` tells the function to fill the values with counts of crimes.

Note the difference from `pivot_longer()` in terms of the arguments which uses `...._to =` whereas `pivot_wider()` uses `..._from =`.

If you are a visual learner, think of it this way:

```{mermaid}
flowchart LR

A["LONG DATA<br/>(values stacked into rows)<br/><pre>
┌───────────────┬──────┬─────────┐
│ neighborhood  │ year │ crimes  │
├───────────────┼──────┼─────────┤
│ A             │ 2020 │ 50      │
│ A             │ 2021 │ 60      │
│ A             │ 2022 │ 30      │
│ B             │ 2020 │ 45      │
│ B             │ 2021 │ 90      │
│ B             │ 2022 │ 40      │
└───────────────┴──────┴─────────┘
</pre>"]

B["pivot_wider()<br/>
names_from = year<br/>
values_from = crimes"]

C["WIDE DATA<br/>(columns contain values)<br/><pre>
┌───────────────┬───────────────┬───────────────┬───────────────┐
│ neighborhood  │ crimes_2020   │ crimes_2021   │ crimes_2022   │
├───────────────┼───────────────┼───────────────┼───────────────┤
│ A             │ 50            │ 60            │ 30            │
│ B             │ 45            │ 90            │ 40            │
└───────────────┴───────────────┴───────────────┴───────────────┘
</pre>"]

A --> B --> C
```

Note that this works if we are using multiple variables as well.
For example, say we were looking at arrests and calls for service in a neighborhood and our **long** data looked like this:

```{r}

data_long <- tibble(
  neighborhood = c("A", "A", "A", "B", "B", "B" ),
  year = c( "2020", "2021", "2022","2020", "2021", "2022" ),
  arrests = c(50, 60, 30, 45, 90, 40 ),
  calls = c( 100, 125, 90, 265, 75, 182 )
)

data_long

```

Then, we can create columns by year for arrests and calls like this:

```{r}

data_wide <- data_long |>
  pivot_wider(
    names_from = year,
    values_from = c( arrests, calls )
  )

data_wide

```

In general, use `pivot_wider()` when you’re making a table for a report, you want side-by-side comparisons, you’re exporting data (Excel, dashboards), and so on.
In comparing the two functions, think of it like this: `pivot_longer()` helps us prepare data for analysis, whereas `pivot_wider()` helps us present results.

## Test Your Knowledge Exercises
