```{r setup, include=FALSE}

knitr::opts_chunk$set(
  echo = TRUE,
  eval = TRUE,
  message=FALSE, 
  warning=FALSE, 
  fig.width = 10,
  fig.align = "center"
  )

library( DWVpack )
library( tidyverse )
library( dplyr )
library( ggplot2 )
library( scales )

```

# Data Visualization with `ggplot2` {.unnumbered}

In the prior chapters we have worked on obtaining an open-source data object, tidying it, and constructing variables.
Now, we want to use those data to visualize the information.
Data visualization transforms raw numbers into meaningful visual patterns that humans can interpret quickly.

Visualization helps analysts:

-   Detect trends over time
-   Compare categories
-   Examine distributions
-   Explore relationships between variables
-   Identify outliers and anomalies
-   Communicate findings effectively

Without visualization, important insights can remain hidden inside tables of data.
Furthermore, the visual display of information reaches a broader audience than tabular display.

For example, consider the following table:

| Month     | Property Crimes |
|:----------|----------------:|
| January   |             120 |
| February  |             150 |
| March     |             170 |
| April     |             220 |
| May       |             245 |
| June      |             260 |
| July      |             290 |
| August    |             310 |
| September |             295 |
| October   |             340 |
| November  |             360 |
| December  |             410 |

The table provides information (property crime each month), but a visualization makes patterns much easier to recognize immediately.

For example, a line graph would quickly reveal: - overall growth throughout the year - periods of rapid increase - slight slowdowns or dips - seasonal patterns - long-term trends

You *can* see these if you look through the table, but the visualization of the data makes them **easier** to see:

```{r, echo=FALSE}

crimes <- data.frame(
  month = c("January", "February", "March", "April",
            "May", "June", "July", "August",
            "September", "October", "November", "December"),
  crimes = c(120, 150, 170, 220,
            245, 260, 290, 310,
            295, 340, 360, 410)
)

ggplot(crimes, aes(x = month, y = crimes, group = 1)) +
  geom_line(color = "firebrick", linewidth = 1) +
  geom_point(size = 3, color = "black") +
  labs(
    title = "Monthly Property Crimes Throughout the Year",
    x = "Month",
    y = "Crimes"
  ) +
  theme_minimal()

```

Visualization improves comprehension by allowing analysts to identify relationships and patterns that may not be obvious in raw numerical tables alone.

## Principles of Effective Visualization

Before we get started with creating visual displays, we should think about what properties are desirable.
Good visualizations are: clear, accurate, simple, and focused on insight.
The reader (viewer?) should be able to look at the visualization and make sense of it without being led astray by too many colors, misleading axes, cluttered labels, unnecessary 3D effects, or overcomplicated graphics.

### Design Principles

Here are several design principles to keep in mind when creating visualizations:

-   Simplicity - Remove unnecessary elements
-   Readability - Labels and titles should be easy to understand
-   Appropriate Chart Selection - Use the correct chart for the question being asked
-   Color with Purpose - Colors should highlight meaning, not distract

For an excellent review of visualization (with some really great examples), check out Kieran Healy's [Look at Data](https://socviz.co/01-look-at-data.html#what-makes-bad-figures-bad) chapter of his [Data Visualization: A Practical Introduction](https://socviz.co/) book.

## What is ggplot2?

`ggplot2` is a data visualization package in R based on the ["Grammar of Graphics"](https://link.springer.com/book/10.1007/0-387-28695-0), a framework for building data visualizations by combining independent visual components together in layers.
Rather than treating each chart type as a completely separate object, the Grammar of Graphics views all graphics as combinations of:\
- data - mappings - geometric shapes - scales - coordinate systems - visual themes

This framework was originally developed by statistician [Leland Wilkinson](https://en.wikipedia.org/wiki/Leland_Wilkinson) and later implemented in R through the `ggplot2` package by [Hadley Wickham](https://hadley.nz/).

The name `ggplot2` literally means: - gg = Grammar of Graphics - plot2 = second version of the plotting system

Traditional graphing systems often require users to memorize separate commands for each chart type.
In contrast, `ggplot2` uses a consistent structure for nearly every visualization.
This means that once you learn the grammar, you can build many different types of graphics while only changing specific layers.

The utility might not be immediately obvious, but as we start building visualizations, it will become more clear (hang with me on this one!) This layered structure makes visualization more logical, flexible, easier to customize, and easier to interpret.

### Key Components

Every `ggplot2` plot has three key components:

-   **data**
-   A set of **aesthetic mappings** between variables in the data and visual properties
-   At least one layer which describes how to render each observation, created with a **geom** function

Here’s a simple example:

```{r}

# define the data
crime_dat <- data.frame(
  month = c("January", "February", "March", "April",
            "May", "June", "July", "August",
            "September", "October", "November", "December"),
  crimes = c(120, 150, 170, 220,
            245, 260, 290, 310,
            295, 340, 360, 410)
)

# plot it
ggplot( crime_dat, 
        aes( x = crimes, y = month ) ) + 
  geom_point()

```

This produces a scatterplot defined by:

-   Data: `crime_dat`
-   Aesthetic mapping: number of crimes, `crimes`, mapped to the x position, and time, `month`, mapped to the y position
-   Layer: points

The dataset contains the variables that will be visualized, the aesthetic mapping determines how variables are connected to visual properties, and layer is how the information is displayed.

Run this code and think about what is happening:

```{r}

ggplot( crime_dat, 
        aes( x = crimes, y = month ) )

```

Why are there no dots?
Let's think about what we did.
We have the data and we have the aesthetic mapping, but we do not have the **layer**.
We add the layer with the `+ geom_point()` line (hence the inclusion of the `+` operator)

This basic point is key to understating the power of the `ggplot2` for visualization.
Instead of creating an entire chart with one complicated command, we start by stipulating our data and mapping the variables to their positions, then we add layers, then we customize the appearance.
This is the basic setup that allows us to develop very sophisticated visual displays of information.

## Crimes in Phoenix

Recall from the prior chapters that we spent some time tidying the `phx_crime` object in `DWVpack`.
We can now use those tidied data (tidier data?) to create a visualization.
Specifically, we want to recreate this plot:

```{r, echo=FALSE}

crime_monthly <- tidy_phx_crime  |> 
  select( year, month ) |> 
  arrange( year, month ) |> 
  group_by( year, month  ) |> 
  summarize( crimes = n() , .groups = "drop" )

ggplot(crime_monthly, aes(x = month_num, y = crimes, group = year, color = factor(year))) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_x_continuous(
    breaks = 1:12,
    labels = month.abb
  ) +
  labs(
    title = "Crimes by Month",
    x = "Month",
    y = "Number of Crimes",
    color = "Year"
  ) +
  theme_minimal()

```

Now, think for a bit about what we need.
Effectively, we want a data object to plot that has a column for years, months, and number of crimes.
We can use `dplyr` to create the data object.
Let's start with using the `arrange()` and `group_by()` functions to sort the data by "year" and then "month" and then we will group them to create a summary measure that is the count of crimes:

```{r, eval=FALSE}

crime_monthly <- tidy_phx_crime  |> 
  select( year, month ) |> 
  arrange( year, month ) |> 
  group_by( year, month  ) |> 
  summarize( crimes = n() , .groups = "drop" )

```

Now, take a look at the data using `View( crime_monthly )` (it will pop up a separate window pane).
Let's walk through this to see what we are doing:

-   `crime_monthly <- tidy_phx_crime` says to create a new object called `crime_monthly` from the `tidy_phx_crime` object. We create a new object since that will be the data just for this plot.
-   `select( year, month )` takes just the `year` and `month` variables.
-   `arrange( year, month )` sorts the data first by `year` and then by `month`. Again, look at how the data are sorted using the `View( crime_monthly )` command.
-   `group_by( year, month  )` groups the data first by `year` and then by `month`.
-   `summarize( crimes = n() , .groups = "drop" )` creates a new column called `crimes` which is a count for each month of each year (because they data are grouped that way). The `.groups = "drop"` command just tells R to remove the grouping from the object after it is done creating the new column.

Now that we have our data object, we can start building our plot.
As we stated above, every `ggplot2` plot has three key components: **data** (which we just created), a set of **aesthetic mappings** between variables in the data and visual properties, and at least one layer which describes how to render each observation, created with a **geom** function.

Let's set up our mapping by making the x-axis the month and the y-axis the count of crimes:

```{r}

ggplot( crime_monthly, aes( x = month, y = crimes, group = year ) )
       
```

Note that the plot is blank because we have not set up the third piece, the layer.

To set up a layer of points, we can use the `geom_point()` function:

```{r}

ggplot( crime_monthly, aes( x = month, y = crimes, group = year ) ) +
  geom_point()
       
```

Ok.
That is a bit better but we cannot see which year is which.
To do this, we can add color using the `color =` argument.
The color aesthetic maps a variable to different colors in the plot:

```{r}

ggplot( crime_monthly, aes( x = month, y = crimes, group = year, color = factor( year ) ) ) +
  geom_point()
       
```

Here, we included `color = factor( year )` to indicate that we want `year` to be converted from numeric years into categories.
This is so that `ggplot` treats them as separate groups (rather than a continuous scale of values).

The points are helpful, but adding lines here would help a great deal.
We can also add lines using the `geom_line()` function.
Note that we are just adding another layer:

```{r}

ggplot( crime_monthly, aes( x = month, y = crimes, group = year, color = factor( year ) ) ) +
  geom_point() +
  geom_line()
       
```

HERE WITH WORKING THROUGH; lets get this cleaned up

HERE IS THE REST OF WHAT TO GO THROUGH

ggplot(crime_monthly, aes(x = month_num, y = crimes, group = year, color = factor(year))) + geom_line(linewidth = 1) + geom_point(size = 2) + scale_x_continuous( breaks = 1:12, labels = month.abb ) + labs( title = "Crimes by Month", x = "Month", y = "Number of Crimes", color = "Year" ) + theme_minimal()

<br>

NEED EXERCISES FOR TEST YOUR KNOWLEDGE
