```{r setup, include=FALSE}

knitr::opts_chunk$set(
  echo = TRUE,
  eval = TRUE,
  message=FALSE, 
  warning=FALSE, 
  fig.width = 10,
  fig.align = "center"
  )

library( DWVpack )
library( tidyverse )
library( dplyr )
library( stringr )
library( lubridate )

```

# Data Transformation {.unnumbered}

When we first start working with data, it’s easy to assume that once we load a dataset into R, we’re ready to analyze it.
In reality, that’s almost never the case.
Most open data is messy, inconsistent, and not immediately useful for answering questions.
It is *wild*.

In the previous chapter we discussed how to get data **tidy**, but even with a tidy dataset we are not yet ready for visualization and analysis.
We still have to get the objects of analysis and visualization prepared.

That’s where **data transformation** comes in.
Data transformation is the process of taking raw data and turning it into something we can actually use for analysis.
We reshape variables, create new ones, and standardize messy information so that patterns can be identified and decisions can be made.

For example, a dataset might give us:

-   A date like “2023-07-14”
-   A time stored as a number like “23”
-   A text field with inconsistent labels like “phx”, “Phoenix”, and “PHOENIX”

On their own, these aren’t very helpful.
But through transformation, we can turn them into: - Day of the week (Was this a weekend incident?)\
- Time category (Did this occur during a night shift?)\
- Cleaned location data (So everything is consistently labeled)

In this chapter, we’re going to focus on three major types of transformations:

-   First, *logical* transformations. These allow us to ask yes/no questions of our data and create new variables based on conditions. For example: Was the incident on a weekend? Did it occur at night? These kinds of variables are essential for analysis.
-   Second, *string* transformations. Many important variables, such as location descriptions or offense types, are stored as text. We’ll learn how to clean, standardize, and search text data so that it becomes usable.
-   Third, date and time transformations. Time is one of the most important dimensions in crime data. We’ll learn how to convert raw date values into meaningful components like month, day of week, and time of day.

All of these transformations have one ultimate goal: to prepare our data for analysis and visualization.

Why is this important?
When we move into visualization using `ggplot2` and start building visual displays, we won’t just be plotting data we pulled from an open source location.
We’ll be plotting variables that we created, like shifts, weekends, trends over time, and categorized outcomes.

## Logical Vectors & Conditional Operations

At its core, a **logical vector** is just a series of `TRUE` and `FALSE` values.
These come from asking questions of your data.

For example:\
- Is the hour greater than 20?\
- Is the day Saturday or Sunday?\
- Does this record match a certain condition?

In R, every time you ask a question like that, you get back `TRUE` or `FALSE`.

To illustrate, let's start by creating a toy data set of crime incidents to work with:

```{r}

dat <- data.frame(
  id = 1:10,
  INC_HOUR = c(2, 9, 14, 21, 23, 7, 12, 18, 3, 16),
  INC_DAY_WEEK = c(
    "Monday", "Tuesday", "Wednesday", "Thursday", "Friday",
    "Saturday", "Sunday", "Monday", "Friday", "Sunday"
  ),
  OFFENSE = c(
    "Theft", "Assault", "Burglary", "Theft", "Robbery",
    "Assault", "Theft", "Burglary", "Robbery", "Assault"
  )
)

dat

```

Let's think of a few questions we can ask with the data:

-   Which incidents occur after 7?
-   Which incidents occur on Friday?

The way we ask questions with logical vectors is to use comparison operators:

-   `==` equal to (note: `=` is used to assign objects in R so we use `==` for an equality statement)
-   `!=` not equal to
-   `>` greater than
-   `<` less than
-   `>=` greater than or equal to
-   `<=` less than or equal to

#### Which incidents occur after 7?

To answer this we can use the `>` operator like so:

```{r, eval = FALSE}

dat$INC_HOUR > 7

```

Let's take a closer look at our code: we identify the object `dat`, we use the `$` sign to indicate the specific column we are interested in, here it is `INC_HOUR`, and then we state the logical statement `> 7`.
Before you run this line, look at the data and think about which cases are going to be `TRUE` and which are going to be `FALSE`.

When you run it, you get this:

```{r, echo=FALSE}

dat$INC_HOUR > 7

```

It is a vector of logical values telling us for which case the condition is `TRUE` or `FALSE`.
For example, the first case has a value of `r dat$INC_HOUR[1]` for the `INC_HOUR` variable so the value for the logical vector is `r dat$INC_HOUR[1] > 7` because it is not greater than 7.

#### Which incidents occur on Friday?

```{r, eval = FALSE}

dat$INC_DAY_WEEK == "Friday"

```

Here, we are saying which cases are equal to `"Friday"`.
This is an important point to pause and remind you that the R programming language is case sensitive.
Also, that `"Friday"` is a `character` (as opposed to say, `numeric`).
Try these and see if it gives you the same results:

```{r, eval = FALSE}

dat$INC_DAY_WEEK == "friday"

dat$INC_DAY_WEEK == "FRIDAY"

dat$INC_DAY_WEEK == "Friday "

```

These (should) come back as all `FALSE`.
*Why?*

These are all `FALSE` because the condition we stated, that `dat$INC_DAY_WEEK` is equal to `"friday"` or `"FRIDAY"` or `"Friday "`, is not met by any case.
This is because our day of the week variable is a `character` so it wants the EXACT string of letters.
Can you see how these three instances are different?

We will discuss strings more below and how to edit them (particularly when there are inconsistencies).

Operators can be used together to satisfy multiple logical conditions using the `|` ("or") or the `&` ("and") operators:

```{r, eval=FALSE}

# or
dat$INC_HOUR >= 7 | dat$INC_DAY_WEEK != "Friday"

# and
dat$INC_HOUR >= 7 & dat$INC_DAY_WEEK != "Friday"

```

Think about what the first statement above says: "incident hour is greater than or equal to 7 **or** incident day is not equal to Friday".
Think about what the second statement above says: "incident hour is greater than or equal to 7 **and** incident day is not equal to Friday".

A final operator that is really useful is the `%in%` operator.
This allows us to specify a set of values or strings and then extract the cases that meet that condition.
For example we could write:

```{r, eval=FALSE}

dat$INC_DAY_WEEK == "Saturday" | dat$INC_DAY_WEEK ==  "Sunday"

```

Or, we could write it like this:

```{r, eval=FALSE}

dat$INC_DAY_WEEK %in% c( "Saturday", "Sunday" )

```

These statements give the same logical vector, but one is cleaner.
For example, what if we have several days:

```{r, eval=FALSE}

# messy
dat$INC_DAY_WEEK == "Friday" | dat$INC_DAY_WEEK == "Saturday" | dat$INC_DAY_WEEK == "Sunday"

# clean
dat$INC_DAY_WEEK %in% c("Friday", "Saturday", "Sunday")

```

As we move on with the material, the use of logical vectors will make a lot more sense because you will see that they are a convenient way to filter data.
For example, in the prior chapters we saw how the `filter()` function in the `dplyr` package could be used to select specific columns.
But, we can plug a logical vector into the `filter()` function to get a subset of data:

```{r}

dat |> 
  dplyr::filter( INC_DAY_WEEK == "Saturday" )

```

Here, we took just those cases meeting the condition of the logical vector (note: I used `dplyr::filter` to call the `dplyr` package rather than loading the library with `library( dplyr )`, to save space).

Suppose we want cases that are after noon and on the weekend:

```{r}

dat |> 
  dplyr::filter( dat$INC_HOUR >= 12 & INC_DAY_WEEK %in% c( "Saturday", "Sunday" ) )

```

## Working with Strings (Text Data)

Now we shift from logical thinking to text data, which are everywhere in open-source datasets.

Think about variables like: - Offense descriptions\
- Location fields - City names\
- Narrative reports

Unlike numbers, text data are often inconsistent ("Phoenix", "PHOENIX", "phx") and messy (extra spaces, abbreviations) making them hard to analyze directly.

Let's create a separate toy dataset to work with for strings.

```{r}

dat <- data.frame(
  id = 1:12,
  city = c(
    "Phoenix", "PHOENIX ", "phx", "Tempe", "tempe", " Glendale",
    "GLENDALE", "Mesa", "mesa ", "Scottsdale", "SCOTTSDALE", "scottsdale"
  ),
  offense = c(
    "Assault", "ASSAULT", "assault ", "Burglary", "BURGLARY", "burglary",
    "Domestic Violence", "domestic violence", "Domestic   Violence",
    "Theft", "THEFT", " theft "
  ),
  location = c(
    "123 MAIN ST", "123 Main St", "123 Main Street",
    "45 E 2ND AVE", "45 E 2nd Ave", "45 e 2ND AVE",
    "700 W RIO SALADO PKWY", "700 W Rio Salado Pkwy", "700 W RIO SALADO PARKWAY",
    "100 N CENTER RD", "100 N Center Rd", "100 N CENTER RD"
  )
)

dat

```

Take a look through the data.
Look for the following issues we commonly find with string variables: - inconsistent capitalization - extra spaces\
- abbreviations - punctuation differences - multi-word labels - alternate spellings for the same thing

This is fairly standard for what you will see with open-source data.
Let's take a look at fixing some of these issues.

### Standardize Case

A very common first step is to make everything the same case.
Why do this?
There are several reasons: It makes values easier to compare, it reduces duplicates caused only by case differences, it makes later cleaning more consistent.

For example, using the `unique()` function, we can list all the unique values of a variable.
Since these are strings, we are looking at the unique strings:

```{r}

unique( dat$city )

```

We can see that there are several different cases for each name.
We can fix this by using some functions from the `stringr` package in conjunction with functions from `dplyr`.
You can install the package using `install.packages( "stringr" )`.

```{r, eval=FALSE}

install.packages( "stringr" )
library( stringr )

library( dplyr )

```

Now that the functions are loaded, let's adjust the cases to be consistent:

```{r}

dat <- dat |>
  mutate(
    city = str_to_title( city ),
    offense = str_to_title( offense ),
    location = str_to_upper( location )
  )

```

We used a few functions, here is what they are doing: -`str_to_title()` converts to title case, where only the first letter of each word is capitalized.
If you look, you can see that the first letter of "city" and "offense" are now capitalized.
- `str_to_upper()` converts to upper case.
For "location" we made all the letters upper case.

There is also: - `str_to_lower()` which converts to lower case.
- `str_to_sentence()` which convert to sentence case, where only the first letter of sentence is capitalized (nice for long strings like quotes or descriptions).

### Remove White Space

Extra spaces are one of the most common hidden problems.
Take a look through the `dat` object and see if you can spot instances where there is extra space (i.e. white space) that should not be there.

For the variable "city" you can see that the second and ninth cases have an extra space:

```{r}

dat$city

```

To fix this, we can use the `str_squish()` function which removes leading spaces (space in front of a word), removes trailing spaces (space after a word), and collapses multiple spaces into one:

```{r}

dat <- dat |>
  mutate(
    city = str_squish( city ),
    offense = str_squish( offense ),
    location = str_squish( location )
  )

```

This looks much better, we can see the difference by examining the unique strings for each variable again:

```{r}

unique( dat$city )

unique( dat$offense )

unique( dat$location )

```

### Group Similar Values

As you can see above, we still have the issue that different naming and abbreviations are used for the same city.
We can fix this by building on what we saw above with the `%in%` operator and the `case_when()` function.
Essentially what we want to do is find all the unique ways that a city is named, and replaced those with a single consistent name.

To do this, we will create a new variable called `city_clean` which is the cleaned version of the names.
We will create a new variable because we don't want to write over the old names (in case we want them for reference).

```{r}

dat <- dat |>
  mutate(
    city_clean = case_when(
      city %in% c( "Phx", "Phoenix" ) ~ "Phoenix",
      city %in% c( "Tempe" ) ~ "Tempe",
      city %in% c( "Glendale" ) ~ "Glendale",
      city %in% c( "Mesa" ) ~ "Mesa",
      city %in% c( "Scottsdale" ) ~ "Scottsdale",
      TRUE ~ city
    )
  )

```

Ok, that was a lot so let's unpack it: - we are using `mutate` to create a new variable.
- we are using `case_when()` to check conditions we specify from the first to the last value of the variable "city".
What the function is doing is taking each case and if the condition is TRUE, it assigns the value on the right side.
- the statement `city %in% c("Phx", "Phoenix") ~ "Phoenix"` says "take the variable city and if the string is "Phx" or "Phoenix" make the string in city_clean be "Phoenix". - we repeated this for Temple, Glendale, Mesa, and Scottsdale. - the last line `TRUE ~ city` tells the function what to do if the conditions above are not met. For example, what should it do if not of the conditions are above are met? Well, we want it to return the value for "city" (otherwise it will return NA; try it!)

### Replace Text

Sometimes you do not want a new label, you just want to fix the text.
We can use the `str_replace_all()` function to accomplish this task.

For example, take a look at the "location" variable and you will see that there are a few different usages of *street*.
If we want these to be consistent, then we can replace the text:

```{r}

dat <- dat |>
  mutate(
    location = str_replace_all( location, "STREET", "ST" ),
    location = str_replace_all( location, "PARKWAY", "PKWY" ),
    location = str_replace_all( location, "ROAD", "RD" )
  )

```

### Putting It All Together

We have discussed these separately, but once we know what we want to clean up we can put it all together in a single workflow.

Since we have been writing over our `dat` object, we need to recreate it from the original:

```{r}

dat <- data.frame(
  id = 1:12,
  city = c(
    "Phoenix", "PHOENIX ", "phx", "Tempe", "tempe", " Glendale",
    "GLENDALE", "Mesa", "mesa ", "Scottsdale", "SCOTTSDALE", "scottsdale"
  ),
  offense = c(
    "Assault", "ASSAULT", "assault ", "Burglary", "BURGLARY", "burglary",
    "Domestic Violence", "domestic violence", "Domestic   Violence",
    "Theft", "THEFT", " theft "
  ),
  location = c(
    "123 MAIN ST", "123 Main St", "123 Main Street",
    "45 E 2ND AVE", "45 E 2nd Ave", "45 e 2ND AVE",
    "700 W RIO SALADO PKWY", "700 W Rio Salado Pkwy", "700 W RIO SALADO PARKWAY",
    "100 N CENTER RD", "100 N Center Rd", "100 N CENTER RD"
  )
)

dat

```

Now, let's build out the full workflow:

```{r}

dat_clean <- dat |>
  
  # fix the case
  mutate(
    city = str_to_title( city ),
    offense = str_to_title( offense ),
    location = str_to_upper( location )
  ) |> 
  
  # clean the white space
  mutate(
    city = str_squish( city ),
    offense = str_squish( offense ),
    location = str_squish( location ) 
    ) |> 
  
  # make names consistent
  mutate(
    city_clean = case_when(
      city %in% c( "Phx", "Phoenix" ) ~ "Phoenix",
      city %in% c( "Tempe" ) ~ "Tempe",
      city %in% c( "Glendale" ) ~ "Glendale",
      city %in% c( "Mesa" ) ~ "Mesa",
      city %in% c( "Scottsdale" ) ~ "Scottsdale",
      TRUE ~ city )
    ) |> 
  
  # replace text for consistency
  mutate(
    location = str_replace_all( location, "STREET", "ST" ),
    location = str_replace_all( location, "PARKWAY", "PKWY" ),
    location = str_replace_all( location, "ROAD", "RD" )
  )

```

## Dates

Dates are notoriously messy and frustrating to handle.
For example, take a look at this fictitious (though representative!) data:

```{r, echo=FALSE}

dat_dates <- data.frame(
  id = 1:10,
  
  incident_date = c(
    "2022-01-15",
    "01/22/2021",
    "March 3, 2024",
    "2023/04/10",
    "5-12-2024",
    "2020-06-18",
    "07/04/2024",
    "August 9 2021",
    "2024.09.14",
    "10/31/2022"
  )
)

dat_dates

```

Do you see the problem?
Everything is in a different format and is inconsistent.
*Wild* data!

Fortunately, there is an excellent package for working with dates and times: `lubridate`.

```{r, eval=FALSE}

install.packages( "lubridate" )

library( lubridate )

```

To see the functionality, let's create our toy data set to work with:

```{r, eval=FALSE}

dat_dates <- data.frame(
  id = 1:10,
  
  incident_date = c(
    "2022-01-15",
    "01/22/2021",
    "March 3, 2024",
    "2023/04/10",
    "5-12-2024",
    "2020-06-18",
    "07/04/2024",
    "August 9 2021",
    "2024.09.14",
    "10/31/2022"
  )
)

dat_dates

```

### The `parse_date_time()` Function

Essentially what we want is the dates to be the same format.
In `lubridate` there is a function, `parse_date_time()`, that does exactly this: it takes the input and parses it into a consistent format.

I will give you the code chunk first, then we will work through it:

```{r}

dat_dates |>
  mutate(
    date_clean = parse_date_time(
      incident_date,
      orders = c( "ymd", "mdy", "B d Y" )
    )
  )

```

This prints out the messy data as well as a new column "date_clean" that has been parsed.
Ok, let's take a look at what the function is doing:\
- using `mutate()`, we are creating a new variable called "date_clean" with the `parse_date_time()` function.
- the `parse_date_time()` function takes the messy data, "incident_date", and then uses some additional info to parse it.
- the `orders =` argument takes several pieces: `ymd` is year-month-day, `mdy` is month-day-year, and `B d Y` is full month name, day, year.

The `orders =` argument might be a bit confusing.
Essentially we are telling the function the different ways that the date is presented in the "incident_date" variable.
For example, in the `dat_dates` object, look at the first case and you will see that the date looks like `r dat_dates[1,2]`.
This is year, month, day format which the `orders =` argument recognizes with the `ymd` value.
How about `mdy`?
Take a look at the second case: `r dat_dates[2,2]`.
And so on.
You can see these different format codes by examining the help page for the `parse_date_time()` function with: `?parse_date_time`.

Working with dates is often a trial and error process of getting a sense of what formats are being used and how to parse them correctly.
For example, let's say we didn't see that some of the dates are in `ymd` format and exclude this from `orders =`:

```{r}

dat_dates |>
  mutate(
    date_clean = parse_date_time(
      incident_date,
      orders = c( "mdy", "B d Y" )
    )
  )

```

We get an error that says it failed to parse several cases.
These cases have a value of `<NA>` in the output object.
If we take a look, then we can see that these are cases with the year-month-day format.
We just need to add that to our `orders =` statement.

Date parsing is one of the messiest parts of open-source data wrangling.
Just keep in mind that the goal is not to memorize formats, but to learn how to recognize patterns and standardize them.

#### Transformation

Let's create an object with the cleaned dates:

```{r}

dat_dates_clean <- dat_dates |>
  mutate(
    date_clean = parse_date_time(
      incident_date,
      orders = c( "ymd", "mdy", "B d Y" )
    )
  )

```

Now that our date variable is clean, we can actually use it to make some other useful variables:

```{r}

dat_dates_clean <- dat_dates_clean |>
  mutate(
    year = year( date_clean ),
    month = month( date_clean, label = TRUE),
    day = day( date_clean )
  )

```

In the statement above we use `mutate()` to create several variables: "year", "month", and "day".
When creating "month" we see the argument `label =` to `TRUE` to keep the labels for the months (otherwise it defaults to a numeric value such as "1" for "Jan").

Now that are data are cleaned, we can use `dplyr` to answer questions like: Which month has the most incidents?

```{r}

dat_dates_clean |>
  group_by( month ) |>  
  summarize( n() )

```

## Example

Let's pull this together by working with an example.
As we have seen in prior chapters, the `DWVpack` package contains a data file on crimes in Phoenix, AZ called `phx_crime`.
Take a look at the `time_date` variable for that object:

```{r}

# load the library
library( DWVpack )

# print out the first few cases using the head() function
head( phx_crime$time_date )

```

We can see that the `time_date` variable contains the **date** of the incident as well as the **time** and looks like this: `r head( phx_crime$time_date )[1]`.
There is a lot of information here.
What we want to do is pull it apart and get several variables:\
- Year - Month - Day - Time

We can use `dplyr` and `lubridate` to help us create these variables.
Let's start by getting the date and time separated.
If you take a look, `r head( phx_crime$time_date )[1]`, you will see that the format is: "month/day/year hour:minute".
There is a function in `lubridate` called `mdy_hm()` that is build specifically for this format and converts a character string into a true date-time object that R can recognize and work with.
For example, `mdy_hm( "11/1/2015 14:30" )` becomes something like `"2015-11-01 14:30:00 UTC"` where `2015-11-01` is the date stored in standardized YYYY-MM-DD format, `14:30:00` is the time in 24-hour format, and `UTC` is the time zone (“Coordinated Universal Time”).

Take a look using the Phoenix data:

```{r}

# create the object
datetime <- mdy_hm( phx_crime$time_date )

# take a look at what it creates
tail( datetime )

```

Next we want to get the date and time as separate objects.
We do this using the `as_date()` and `format()` functions.

```{r}

# create the objects
date <- as_date( datetime )
time <- format( datetime, "%H:%M" )

# inspect the first few cases

tail( date )
tail( time )

```

The formatting codes in `format()` indicate that we want time in 24-hour (`%H`) and minutes (`%M`).

As a final step, we want to pull out calendar components (i.e. year, month, day) as we did above:

```{r}

year  <- year( datetime )

# here, we want to keep the label and don't want it abbreviated
month <- month( datetime, label = TRUE, abbr = FALSE)

day   <- day( datetime )

```

I have presented this in separate steps, but we can put this all together as a "data recipe" using the `mutate()` function:

```{r, eval=FALSE}

phx_crime_clean_date_time <- phx_crime |>
  mutate(
    
    # Parse datetime
    datetime = mdy_hm( time_date ),

    # Separate date and time
    date = as_date( datetime ),
    time = format( datetime, "%H:%M" ),

    # Create components
    year  = year( datetime ),
    month = month( datetime, label = TRUE, abbr = FALSE ),
    day   = day( datetime )

  )

```

We will work with these data in subsequent chapters.
To avoid having to create a tidy data set with the cleaned date variables, `DWVpack` includes a data object called `tidy_phx_crime` which has already cleaned up the date and time variables.

<br>

NEED EXERCISES FOR TEST YOUR KNOWLEDGE
